---
title: Structural Equation Modeling
author: P.11 - Multilevel Model for Change (Part 2)
date: 29.11.2021
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: /Users/mihai/OneDrive/Library/Literature/Library.bib
link-citations: yes
csl: /Users/mihai/OneDrive/Library/Literature/Zotero/styles/apa.csl
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
  - \hypersetup{colorlinks, linkcolor={red!50!black}, citecolor={blue!50!black}, urlcolor={blue!80!black}}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Store the path.
path <- "/Users/mihai/OneDrive/School/Teaching/Structural Equation Modeling - ReMa - Tilburg/2021/Lecture 11 - Multilevel Model for Change (Part 2)/Practical"

# Set the path.
setwd(path)
```

---

# Lab Description

In this assignment you are going to estimate several multilevel models that
reproduce the findings discussed in *lecture 11*. Compare your results with the
findings reported in the lecture slides. Try to use the lecture slides as a
guide through the `R` output.

For this practical you will need the following packages: `lme4`, `ggplot2`, and
`psych`. You can install and load these packages using the following
code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lme4", "ggplot2", "psych"))

# Load the packages.
library(lme4)
library(ggplot2)
library(psych)
```

## Questions

Start by loading the `alcohol.csv` data in `R`, then compute basic descriptive
statistics. The data is available on Canvas in the module corresponding to the
current lab session.

1. Estimate the the *unconditional means model* (i.e., as `model_a`). In this
   model, the variable `alcuse` (i.e., alcohol use) is the dependent variable,
   which is only predicted by the intercept.

    - *Tip*. Recall how interecepts are modeled in simple linear regression, and
      how to allow for the intercepts to vary across individuals.

2. Calculate the *interclass correlation coefficient* (ICC) from `model_a`.

3. Estimate the *unconditional growth model* (i.e., as `model_b`). In this
   model, allow for random variation in the `age_14` variable, which captures
   the effect of time.

    - *Note.* The variable `age_14` by subtracting $14$ from the variable `age`.
      Therefore, variable `age_14` holds $0$ for age $14$, $1$ for age $15$, and
      $2$ for age $16$.

4. Estimate another model (i.e., `model_c`), where the variable `coa` predicts
   both the initial status and the rate of change in variable `alcuse`.

    - *Note.* The variable `coa` refers to whether the children belongs to a
      family with an alcoholic parent, i.e., coded as $1$, and $0$ otherwise.

5. Calculate the proportional reduction in variance in the initial status and
   the rate of change due to including the `coa` predictor in the model.

6. Estimate another (i.e., `model_d`) in which the variable `peer` is added to
   `model_c` to explain the initial status and the rate of change in `alcuse`.

    - *Note.* The variable `peer` is a measure of peer alcohol use.

7. Calculate the proportional reduction in variance in the initial status and
   the rate of change due to including the `peer` predictor in the model.

8. Estimate another model (i.e., `model_e`), in which the non-significant effect
   of varaible `coa` on the rate of change is removed.

9. Estimate another model (i.e., `model_f`) based on `model_e`, but with
   intercepts that describe a child of non-alcoholic parents with an average
   value of `peer` (i.e., use the centered variable `cpeer`).

10.	Perform a *Likelihood-Ratio Test* (LRT) in which you simultaneously compare
    `model_c`, `model_d`, and `model_e`. What do you conclude?
