---
title: Structural Equation Modeling
author: P.08 - MIMIC Models and Instrumental Variables
date: "`r format(Sys.time(), '%B %d, %Y (%H:%M:%S)')`"
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: ../libs/literature.bib
csl: ../libs/apa.csl
link-citations: yes
colorlinks: true
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Set the path only for interactive sessions.
if (interactive()) setwd("./Practical 08 - MIMIC Models and Instrumental Variables")

# Load libraries.
library(lavaan)
library(semPlot)
```

---

# Lab Description

For this practical you will need the following packages: `lavaan` and `semPlot`.
You can install and load these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot", "mvtnorm", "GGally"))

# Load the packages.
library(lavaan)
library(semPlot)
```

## Exercise 1

Estimate the model in *Figure 1* in `lavaan` and examine if there is evidence of
Differential Item Functioning (DIF) in the measurement instruments. To help you
get started, you are provided with the code that contains the correlations and
standard deviations corresponding to the model depicted in *Figure 1*.

![Reproduction of Figure 7.5 from @brownConfirmatoryFactorAnalysis2014 [p. 275]](./assets/ex_1_figure_1.jpg){width=100%}

Standard deviations and correlations.

```{r size='scriptsize'}
# Standard deviations.
sd <- "2.26 2.73 2.11 2.32 2.61 2.44 0.50"

# Correlations.
cor <- "
    1.000
    0.705   1.000
    0.724   0.646   1.000
    0.213   0.195   0.190   1.000
    0.149   0.142   0.128   0.521   1.000
    0.155   0.162   0.135   0.557   0.479   1.000
    -0.019  -0.024  -0.029  -0.110  -0.074  -0.291   1.000
"

# Get covariances.
cov <- getCov(cor, sds = sd, names = c("S1", "S2", "S3", "A1", "A2", "A3", "sex"))
```

\newpage

# References
