---
title: Structural Equation Modeling
author: P.08 - MIMIC Models and Instrumental Variables
date: "`r format(Sys.time(), '%B %d, %Y (%H:%M:%S)')`"
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: ../libs/literature.bib
csl: ../libs/apa.csl
link-citations: yes
colorlinks: true
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Set the path only for interactive sessions.
if (interactive()) setwd("./Practical 08 - MIMIC Models and Instrumental Variables")

# Load data for exercise two.
data_ex_2 <- read.csv("./data/card.csv")

# Load libraries.
library(lavaan)
library(semPlot)
```

---

# Lab Description

For this practical you will need the following packages: `lavaan` and `semPlot`.
You can install and load these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot", "mvtnorm", "GGally"))

# Load the packages.
library(lavaan)
library(semPlot)
```

## Exercise 1

Estimate the model in *Figure 1* in `lavaan` and examine if there is evidence of
Differential Item Functioning (DIF) in the measurement instruments. To help you
get started, you are provided with the code that contains the correlations and
standard deviations corresponding to the model depicted in *Figure 1*.

![Reproduction of Figure 7.5 from @brownConfirmatoryFactorAnalysis2014 [p. 275]](./assets/ex_1_figure_1.jpg){width=100%}

Standard deviations and correlations.

```{r size='scriptsize'}
# Standard deviations.
sd <- "2.26 2.73 2.11 2.32 2.61 2.44 0.50"

# Correlations.
cor <- "
    1.000
    0.705   1.000
    0.724   0.646   1.000
    0.213   0.195   0.190   1.000
    0.149   0.142   0.128   0.521   1.000
    0.155   0.162   0.135   0.557   0.479   1.000
    -0.019  -0.024  -0.029  -0.110  -0.074  -0.291   1.000
"

# Get covariances.
cov <- getCov(cor, sds = sd, names = c("S1", "S2", "S3", "A1", "A2", "A3", "sex"))
```

\newpage

We start by specifying the syntax for the *MIMIC* model.

```{r size="footnotesize", dev="pdf", fig.width=20, fig.height=9, out.width="450px", fig.align="center"}
# Model syntax.
model_ex_1 <- "
    # Measurement part.
    social  =~ S1 + S2 + S3
    agoraph =~ A1 + A2 + A3

    # Regression equations.
    social ~ sex
    agoraph ~ sex
    A3 ~ sex

    # Covariances.
    social ~~ agoraph
"

# Fit the model.
model_ex_1_fit <- cfa(model_ex_1, sample.cov = cov, sample.nobs = 730)

# Visualize the model.
semPaths(model_ex_1_fit, what = "paths", whatLabels = "est")

# Model summary.
summary(model_ex_1_fit, fit.measures = TRUE, standardized = TRUE, modindices = TRUE)
```

The *MIMIC* model provides a good fit to the data, with a $\chi^2(11) = 3.80$,
$p\text{-value} = .98$, $RMSEA = 0.00$, and  $CFI = 1.00$.

Regarding the evidence for DIF, the following paragraph from
@brownConfirmatoryFactorAnalysis2014 [p. 280] is relevant:

> Consistent with the researcher’s predictions, the results of the *MIMIC* model
> show that the A3 indicator is not invariant for males and females (akin to
> intercept non-invariance in multiple-groups CFA). This is reflected by the
> significant direct effect of `sex` on the `A3` indicator ($z = 6.65$, $p <
> .001$) that is not mediated by `agoraphobia`. In other words, when the latent
> variable of `agoraphobia` is held constant, there is a significant direct
> effect of `sex` on the `A3` indicator. Thus, at any given value of the factor,
> women score significantly higher on the `A3` indicator than men (by $.985$
> units, or nearly a full point on the $0$–$8$ scale). This is evidence of
> *differential item functioning*; that is, the item behaves differently as an
> indicator of `agoraphobia` in men and women.

## Exercise 2

Open the dataset `card.csv` available on Canvas in the folder corresponding to
the current practical. This dataset contains several variables used by David
Card (1995) to estimate the causal effect of education on wages using proximity
to college as an instrumental variable. You can find more information about this
dataset at [this link](https://eml.berkeley.edu/~card/papers/geo_var_schooling.pdf).

Set the working directory to the location where your data file has been
downloaded and load the data.

```{r size="footnotesize", eval=FALSE}
# For example.
setwd("/Users/mihai/Downloads")

# Load data.
data_ex_2 <- read.csv("card.csv")

# Inspect the data.
View(data_ex_2)
```

a. Estimate a model in which you only regress `lwage` on `educ` (i.e., without
   including an instrumental variable). What do you conclude from this
   regression?

```{r size="footnotesize", dev="pdf", fig.width=9, fig.height=3, out.width="450px", fig.align="center"}
# Model syntax.
model_ex_2_a <- "lwage ~ educ"

# Fit the model.
model_ex_2_a_fit <- sem(model_ex_2_a, data_ex_2)

# Visualize the model.
semPaths(model_ex_2_a_fit, what = "paths", whatLabels = "est")

# Model summary.
summary(model_ex_2_a_fit, standardized = TRUE, rsquare = TRUE)
```

b. Re-estimate the model at point *(a)*, but this time with the following
   control variables added: `exper`, `expersq`, `black`, `south`, `fatheduc`,
   and `motheduc`.

```{r size="footnotesize", dev="pdf", fig.width=20, fig.height=9, out.width="450px", fig.align="center"}
model_ex_2_b <- "
    lwage ~ educ + exper + expersq + black + south + fatheduc + motheduc
"

# Fit the model.
model_ex_2_b_fit <- sem(model_ex_2_b, data_ex_2)

# Visualize the model.
semPaths(model_ex_2_b_fit, what = "paths", whatLabels = "est")

# Model summary.
summary(model_ex_2_b_fit, standardized = TRUE, rsquare = TRUE)
```

The problem with treating the direct association between `educ` and `lwage` as a
causal effect is that there are likely many omitted variables that affect both
education and wages. We could control for those variables by measuring them and
including them in the model (i.e., as we did at point *b*). But there is no way
we can control for all possible confounding variables, especially because some
variables are difficult to measure (e.g., ability). It is therefore likely that
education is correlated with the error term in the regression (i.e., a form of
endogeneity), and that our regression coefficient is, in turn, biased to an
unknown degree. David Card proposed to solve this problem by introducing
proximity to college as an instrumental variable. Specifically, `nearc4` was a
dummy indicator variable for whether or not the person was raised in a local
labor market that included a four-year college.

c. Re-estimate the model at point *(b)* with the following additions:
    - add `nearc4` as an instrumental variable for for `educ`, while
      controlling for `fatheduc` and `motheduc`
    - add a covariance between the error terms of `educ` and `lwage`

Does this model provide evidence of endogeneity of educ? Why (not)?

```{r size="footnotesize", dev="pdf", fig.width=20, fig.height=9, out.width="450px", fig.align="center"}
model_ex_2_c <- "
    lwage ~ educ + exper + expersq + black + south + fatheduc + motheduc
    educ ~ nearc4 + fatheduc + motheduc
    lwage ~~ educ
"

# Fit the model.
model_ex_2_c_fit <- sem(model_ex_2_c, data_ex_2)

# Model summary.
summary(model_ex_2_c_fit, standardized = TRUE, rsquare = TRUE)
```

d. Evaluate whether `nearc4` is a weak or strong instrument for dealing with the
   endogeneity of the variable `educ`. Specifically, consider the criteria that a
   strong instrument must meet in order to adequately correct for endogeneity.

```{r size="footnotesize", dev="pdf", fig.width=20, fig.height=9, out.width="450px", fig.align="center"}
model_ex_2_d <- "
    lwage ~ educ + exper + expersq + black + south + fatheduc + motheduc + nearc4
    educ ~ nearc4 + fatheduc + motheduc
"

# Fit the model.
model_ex_2_d_fit <- sem(model_ex_2_d, data_ex_2)

# Visualize the model.
semPaths(model_ex_2_d_fit, what = "paths", whatLabels = "est")

# Model summary.
summary(model_ex_2_d_fit, standardized = TRUE, rsquare = TRUE)
```

Overall, `nearc4` does not appear to be a particularly strong instrumental
variable!

\newpage

# References
