---
title: Structural Equation Modeling
author: P.08 - MIMIC Models and Instrumental Variables
date: 17.11.2021
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: /Users/mihai/OneDrive/Library/Literature/Library.bib
link-citations: yes
csl: /Users/mihai/OneDrive/Library/Literature/Zotero/styles/apa.csl
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
  - \hypersetup{colorlinks, linkcolor={red!50!black}, citecolor={blue!50!black}, urlcolor={blue!80!black}}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Store the path.
path <- "/Users/mihai/OneDrive/School/Teaching/Structural Equation Modeling - ReMa - Tilburg/2021/Lecture 08 - MIMIC Models and Instrumental Variables/Practical"

# Set the path.
setwd(path)

# Load libraries.
library(lavaan)
library(semPlot)
```

---

# Lab Description

For this practical you will need the following packages: `lavaan` and `semPlot`.
You can install and load these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot", "mvtnorm", "GGally"))

# Load the packages.
library(lavaan)
library(semPlot)
```

## Exercise 1

Estimate the model in *Figure 1* in `lavaan` and examine if there is evidence of
Differential Item Functioning (DIF) in the measurement instruments. To help you
get started, you are provided with the code that contains the correlations and
standard deviations corresponding to the model depicted in *Figure 1*.

![Reproduction of Figure 7.5 from @brownConfirmatoryFactorAnalysis2014 [p. 275]](./assets/ex_1_figure_1.jpg){width=100%}

Standard deviations and correlations.

```{r size='scriptsize'}
# Standard deviations.
sd <- "2.26 2.73 2.11 2.32 2.61 2.44 0.50"

# Correlations.
cor <- "
    1.000
    0.705   1.000
    0.724   0.646   1.000
    0.213   0.195   0.190   1.000
    0.149   0.142   0.128   0.521   1.000
    0.155   0.162   0.135   0.557   0.479   1.000
    -0.019  -0.024  -0.029  -0.110  -0.074  -0.291   1.000
"

# Get covariances.
cov <- getCov(cor, sds = sd, names = c("S1", "S2", "S3", "A1", "A2", "A3", "sex"))
```

## Exercise 2

Open the dataset `card.csv` available on Canvas in the folder corresponding to
the current practical. This dataset contains several variables used by David
Card (1995) to estimate the causal effect of education on wages using proximity
to college as an instrumental variable. You can find more information about this
dataset at [this link](https://eml.berkeley.edu/~card/papers/geo_var_schooling.pdf).

a. Estimate a model in which you only regress `lwage` on `educ` (i.e., without
   including an instrumental variable). What do you conclude from this
   regression?

b. Re-estimate the model at point *(a)*, but this time with the following
   control variables added: `exper`, `expersq`, `black`, `south`, `fatheduc`,
   and `motheduc`.

The problem with treating the direct association between `educ` and `lwage` as a
causal effect is that there are likely many omitted variables that affect both
education and wages. We could control for those variables by measuring them and
including them in the model (i.e., as we did at point *b*). But there is no way
we can control for all possible confounding variables, especially because some
variables are difficult to measure (e.g., ability). It is therefore likely that
education is correlated with the error term in the regression (i.e., a form of
endogeneity), and that our regression coefficient is, in turn, biased to an
unknown degree. David Card proposed to solve this problem by introducing
proximity to college as an instrumental variable. Specifically, `nearc4` was a
dummy indicator variable for whether or not the person was raised in a local
labor market that included a four-year college.

c. Re-estimate the model at point *(b)* with the following additions:
    - add `nearc4` as an instrumental variable for for `educ`, while
      controlling for `fatheduc` and `motheduc`
    - add a covariance between the error terms of `educ` and `lwage`

Does this model provide evidence of endogeneity of educ? Why (not)?

d. Evaluate whether `nearc4` is a weak or strong instrument for dealing with the
   endogeneity of the variable `educ`. Specifically, consider the criteria that a
   strong instrument must meet in order to adequately correct for endogeneity.

\newpage

# References
