---
title: Structural Equation Modeling
author: P.08 - MIMIC Models
date: "`r format(Sys.time(), '%B %d, %Y (%H:%M:%S)')`"
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: ../libs/literature.bib
csl: ../libs/apa.csl
link-citations: yes
colorlinks: true
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Set the path only for interactive sessions.
if (interactive()) setwd("./Practical 08 - MIMIC Models")

# Load data for exercise two.
data_ex_2 <- read.csv("./data/card.csv")

# Load libraries.
library(lavaan)
library(semPlot)
```

---

# Lab Description

For this practical you will need the following packages: `lavaan` and `semPlot`.
You can install and load these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot", "mvtnorm", "GGally"))

# Load the packages.
library(lavaan)
library(semPlot)
```

## Exercise 1

Estimate the model in *Figure 1* in `lavaan` and examine if there is evidence of
Differential Item Functioning (DIF) in the measurement instruments. To help you
get started, you are provided with the code that contains the correlations and
standard deviations corresponding to the model depicted in *Figure 1*.

![Reproduction of Figure 7.5 from @brownConfirmatoryFactorAnalysis2014 [p. 275]](./assets/ex_1_figure_1.jpg){width=100%}

Standard deviations and correlations.

```{r size='scriptsize'}
# Standard deviations.
sd <- "2.26 2.73 2.11 2.32 2.61 2.44 0.50"

# Correlations.
cor <- "
    1.000
    0.705   1.000
    0.724   0.646   1.000
    0.213   0.195   0.190   1.000
    0.149   0.142   0.128   0.521   1.000
    0.155   0.162   0.135   0.557   0.479   1.000
    -0.019  -0.024  -0.029  -0.110  -0.074  -0.291   1.000
"

# Get covariances.
cov <- getCov(cor, sds = sd, names = c("S1", "S2", "S3", "A1", "A2", "A3", "sex"))
```

\newpage

We start by specifying the syntax for the *MIMIC* model.

```{r size="footnotesize", dev="pdf", fig.width=20, fig.height=9, out.width="450px", fig.align="center"}
# Model syntax.
model_ex_1 <- "
    # Measurement part.
    social  =~ S1 + S2 + S3
    agoraph =~ A1 + A2 + A3

    # Regression equations.
    social ~ sex
    agoraph ~ sex
    A3 ~ sex

    # Covariances.
    social ~~ agoraph
"

# Fit the model.
model_ex_1_fit <- cfa(model_ex_1, sample.cov = cov, sample.nobs = 730)

# Visualize the model.
semPaths(model_ex_1_fit, what = "paths", whatLabels = "est")

# Model summary.
summary(model_ex_1_fit, fit.measures = TRUE, standardized = TRUE, modindices = TRUE)
```

The *MIMIC* model provides a good fit to the data, with a $\chi^2(11) = 3.80$,
$p\text{-value} = .98$, $RMSEA = 0.00$, and  $CFI = 1.00$.

Regarding the evidence for DIF, the following paragraph from
@brownConfirmatoryFactorAnalysis2014 [p. 280] is relevant:

> Consistent with the researcher’s predictions, the results of the *MIMIC* model
> show that the A3 indicator is not invariant for males and females (akin to
> intercept non-invariance in multiple-groups CFA). This is reflected by the
> significant direct effect of `sex` on the `A3` indicator ($z = 6.65$, $p <
> .001$) that is not mediated by `agoraphobia`. In other words, when the latent
> variable of `agoraphobia` is held constant, there is a significant direct
> effect of `sex` on the `A3` indicator. Thus, at any given value of the factor,
> women score significantly higher on the `A3` indicator than men (by $.985$
> units, or nearly a full point on the $0$–$8$ scale). This is evidence of
> *differential item functioning*; that is, the item behaves differently as an
> indicator of `agoraphobia` in men and women.

\newpage

# References
