---
title: Structural Equation Modeling
author: P.12 - Growth Curve Analysis
date: "`r format(Sys.time(), '%B %d, %Y (%H:%M:%S)')`"
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: ../libs/literature.bib
csl: ../libs/apa.csl
link-citations: yes
colorlinks: true
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Set the path only for interactive sessions.
if (interactive()) setwd("./Practical 12 - Growth Curve Analysis")
```

---

# Lab Description

In this assignment you are going to learn how to estimate a *Latent Growth
Curve* (LGC) model in `R` using the `lavaan`. In the previous lab you learn how
you can use hierarchical models (i.e., multi-level models for nested data). In
this lab, we are are still going to work with nested data (i.e., observations
nested under participants), but this time we are going to approach the problem
from the perspective of *Structural Equation Models* (SEM) For this practical
you will need the following packages: `lavaan` and `semPlot`. You can install
and load these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot"))

# Load the packages.
library(lavaan)
library(semPlot)
```

## Exercise 1

In this exercise, you are going to estimate a LGC model similar as the one
depicted in *Figure 1*, but with six waves instead of four. Specifically, you
are going to investigate changes in body weight over $12$ years (i.e., six waves
of data separated by two years each) using the health and aging data set
`health.dat` with $N = 5335$. Body weight was operationalized as the *Body Mass
Index* (BMI), which is a ratio of weight to square of height (i.e., $\text{kg} /
\text{m}^2$). The data set `health.dat` is available on Canvas in the module
corresponding to this lab. You can use the code below to load the data and set
the variables names.

![Example of a LGT model reproduced from @newsomLongitudinalStructuralEquation2015 [p. 174].](./assets/ex_1_figure_1.png){width=80%}

Set the working directory to the location where your data file has been
downloaded and load the data.

```{r size="footnotesize", eval=FALSE}
# For example.
setwd("/Users/mihai/Downloads")

# Load data.
data <- read.table("health.dat")

# Inspect the data.
View(data)
```

Set the variable names.

```{r size="footnotesize", eval=FALSE}
# Variable names.
variable_names_ex_1 = c(
    "age", "srh1", "srh2", "srh3", "srh4", "srh5", "srh6", "bmi1",
    "bmi2", "bmi3", "bmi4", "bmi5", "bmi6", "cesdna1", "cesdpa1", "cesdso1",
    "cesdna2", "cesdpa2", "cesdso2", "cesdna3", "cesdpa3", "cesdso3",
    "cesdna4", "cesdpa4", "cesdso4", "cesdna5", "cesdpa5", "cesdso5",
    "cesdna6", "cesdpa6", "cesdso6", "diab1", "diab2", "diab3 ", "diab4", "diab5", "diab6"
)

# Set the names.
names(data) <- variable_names_ex_1
```

List the variables.

```{r size="footnotesize", eval=FALSE}
# List variables.
str(data)
```

a. Estimate a *LGC* model in which you set each intercept factor loading equal
   to $1$ and the slope factor loadings equal to $0$, $1$, $2$, $3$, $4$, and
   $5$. Do not include correlated measurement residuals in this model. Evaluate
   the fit of this model, and interpret the mean of the latent intercept and
   mean of the latent slope.

b. Re-estimate the model from point *(a)*, but now add auto-correlations among
   adjacent time points of the measurement residuals (e.g., $\varepsilon_1$ with
   $\varepsilon_2$, $\varepsilon_2$ with $\varepsilon_3$, and so on). Evaluate
   the fit of this model and test its fit against the more restricted model
   estimated at point *(a)*.

c. In the model estimated at point *(b)*, inspect the estimate for the
   covariance between the latent intercept and the latent slope. Provide a
   substantive interpretation for this value.

d. Estimate a model that assumes homogeneity of variance of the measurement
   residuals. In this model, remove the correlated measurement residuals so you
   can test this model against the unconstrained model that was estimated at
   point *(a)*. What can you conclude from the comparison of both models, and
   from the model that included the auto-correlations between error terms?

## Exercise 2

In this exercise you are going to keep using the `health.dat` and extend the
model from *Exercise 1* with a time-invariant covariate, namely the age of the
participants.

a. Compute a new variable `age_c` which is the grand mean centered age of
   the participants.
    - *Tip.* You can obtain the sample mean using `mean(data$age)`.

b. Estimate the conditional *LGC* model of *BMI* by regressing the latent
   intercept and the latent slope on `age_c`.

c. Evaluate the fit of this model, and the effect of `age_c` on the latent
   intercept and latent slope.

d. Compute a new binary age variable called `age_b` for implementing the *MIMIC*
   approach discussed during *Lecture 8*, using the following `R` code:

```{r size="footnotesize", eval=FALSE}
data$age_b <- ifelse(data$age <= 65, 0, 1)
```

e. Estimate the *MIMIC* model by regressing the intercept and the slope factors
   on the binary age variable. Interpret the effects of the binary age variable.

f. Test the same hypothesis of age differences in trajectories, using the
   multi-group approach and comparing those under $65$ to those aged $65$ and
   older.

\newpage

# References
