---
title: Structural Equation Modeling
author: P.12 - Growth Curve Analysis
date: "`r format(Sys.time(), '%B %d, %Y (%H:%M:%S)')`"
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: ../libs/literature.bib
csl: ../libs/apa.csl
link-citations: yes
colorlinks: true
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Set the path only for interactive sessions.
if (interactive()) setwd("./Practical 12 - Growth Curve Analysis")

# Load the data.
data <- read.table("./data/health.dat")

# Load libraries.
library(lavaan)
library(semPlot)
```

---

# Lab Description

In this assignment you are going to learn how to estimate a *Latent Growth
Curve* (LGC) model in `R` using the `lavaan`. In the previous lab you learn how
you can use hierarchical models (i.e., multi-level models for nested data). In
this lab, we are are still going to work with nested data (i.e., observations
nested under participants), but this time we are going to approach the problem
from the perspective of *Structural Equation Models* (SEM) For this practical
you will need the following packages: `lavaan` and `semPlot`. You can install
and load these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot"))

# Load the packages.
library(lavaan)
library(semPlot)
```

## Exercise 1

In this exercise, you are going to estimate a LGC model similar as the one
depicted in *Figure 1*, but with six waves instead of four. Specifically, you
are going to investigate changes in body weight over $12$ years (i.e., six waves
of data separated by two years each) using the health and aging data set
`health.dat` with $N = 5335$. Body weight was operationalized as the *Body Mass
Index* (BMI), which is a ratio of weight to square meter of height (i.e.,
$\text{kg} / \text{m}^2$). The data set `health.dat` is available on Canvas in
the module corresponding to this lab. You can use the code below to load the
data and set the variables names.

![Example of a LGT model reproduced from @newsomLongitudinalStructuralEquation2015 [p. 174].](./assets/ex_1_figure_1.png){width=80%}

Set the working directory to the location where your data file has been
downloaded and load the data.

```{r size="footnotesize", eval=FALSE}
# For example.
setwd("/Users/mihai/Downloads")

# Load data.
data <- read.table("health.dat")

# Inspect the data.
View(data)
```

Set the variable names.

```{r size="footnotesize"}
# Variable names.
variable_names_ex_1 = c(
    "age", "srh1", "srh2", "srh3", "srh4", "srh5", "srh6", "bmi1",
    "bmi2", "bmi3", "bmi4", "bmi5", "bmi6", "cesdna1", "cesdpa1", "cesdso1",
    "cesdna2", "cesdpa2", "cesdso2", "cesdna3", "cesdpa3", "cesdso3",
    "cesdna4", "cesdpa4", "cesdso4", "cesdna5", "cesdpa5", "cesdso5",
    "cesdna6", "cesdpa6", "cesdso6", "diab1", "diab2", "diab3 ", "diab4", "diab5", "diab6"
)

# Set the names.
names(data) <- variable_names_ex_1
```

List the variables.

```{r size="footnotesize", eval=FALSE}
# List variables.
str(data)
```

Specify which fit measures we are interested in:

```{r size='scriptsize'}
# Fit indices to print.
fit_indices <- c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", "rmsea.pvalue", "srmr")
```

a. Estimate a *LGC* model in which you set each intercept factor loading equal
   to $1$ and the slope factor loadings equal to $0$, $1$, $2$, $3$, $4$, and
   $5$. Do not include correlated measurement residuals in this model. Evaluate
   the fit of this model, and interpret the mean of the latent intercept and
   mean of the latent slope.

*It is important to note that instead of functions `sem` and `cfa`, this time we
use the function **`growth`** for fitting the models.*

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
# Model syntax.
model_ex_1_a <- "
    # Latent intercept variable.
    int =~ 1 * bmi1 + 1 * bmi2 + 1 * bmi3 + 1 * bmi4 + 1 * bmi5 + 1 * bmi6

    # Latent slope variable.
    slp =~ 0 * bmi1 + 1 * bmi2 + 2 * bmi3 + 3 * bmi4 + 4 * bmi5 + 5 * bmi6

    # Latent variances and covariances.
    int ~~ int
    slp ~~ slp
    int ~~ slp
"

# Fit model.
model_ex_1_a_fit <- growth(model_ex_1_a, data = data)

# Visualize the model.
semPaths(model_ex_1_a_fit, what = "paths", whatLabels = "std", edge.label.position = 0.7)

# Model summary.
summary(model_ex_1_a_fit, standardized = TRUE, rsquare = TRUE)

# Fit measures.
fitMeasures(model_ex_1_a_fit, fit.measures = fit_indices)
```

Note that when we use `rsquare = TRUE` the `lavaan` documentation indicates the
following:

> If `rsquare = TRUE`, the $r^2$ values for the dependent variables in the model
are printed.

b. Re-estimate the model from point *(a)*, but now add auto-correlations among
   adjacent time points of the measurement residuals (e.g., $\varepsilon_1$ with
   $\varepsilon_2$, $\varepsilon_2$ with $\varepsilon_3$, and so on). Evaluate
   the fit of this model and test its fit against the more restricted model
   estimated at point *(a)*.

We can add the adjacent correlated residuals as follows:

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
# Model syntax.
model_ex_1_b <- "
    # Latent intercept variable.
    int =~ 1 * bmi1 + 1 * bmi2 + 1 * bmi3 + 1 * bmi4 + 1 * bmi5 + 1 * bmi6

    # Latent slope variable.
    slp =~ 0 * bmi1 + 1 * bmi2 + 2 * bmi3 + 3 * bmi4 + 4 * bmi5 + 5 * bmi6

    # Latent variances and covariances.
    int ~~ int
    slp ~~ slp
    int ~~ slp

    # Adjacent correlated residuals.
    bmi1 ~~ bmi2
    bmi2 ~~ bmi3
    bmi3 ~~ bmi4
    bmi4 ~~ bmi5
    bmi5 ~~ bmi6
"

# Fit model.
model_ex_1_b_fit <- growth(model_ex_1_b, data = data)

# Visualize the model.
semPaths(model_ex_1_b_fit, what = "paths", whatLabels = "std", edge.label.position = 0.7)

# Model summary.
summary(model_ex_1_b_fit, standardized = TRUE, rsquare = TRUE)

# Fit measures.
fitMeasures(model_ex_1_b_fit, fit.measures = fit_indices)
```

To compare the fit of `model_ex_1_b` to `model_ex_1_a` we can take a look at the
fit measures and also perform a *Likelihood Ratio Test* (LRT).

```{r size="footnotesize"}
# Put all fit measures in a data frame or convenience.
fit_measures_ex_1 <- data.frame(
    model_ex_1_a = fitMeasures(model_ex_1_a_fit, fit.measures = fit_indices),
    model_ex_1_b = fitMeasures(model_ex_1_b_fit, fit.measures = fit_indices)
)

# Print all fit measures rounded to four decimals.
round(fit_measures_ex_1, 4)

# Perform the LRT.
anova(model_ex_1_a_fit, model_ex_1_b_fit)
```

c. In the model estimated at point *(b)*, inspect the estimate for the
   covariance between the latent intercept and the latent slope. Provide a
   substantive interpretation for this value.

We can use the function `parTable` to list all parameters.

```{r size="footnotesize"}
# List parameters.
lavaan::parTable(model_ex_1_b_fit)
```

In our case, the parameter of interest is located at row $15$.

```{r size="footnotesize"}
lavaan::parTable(model_ex_1_b_fit)[15, ]
```

We obtained a covariance between the latent intercept and slope variables of
$.15$, with a $SE = 0.042$.

d. Estimate a model that assumes homogeneity of variance of the measurement
   residuals. In this model, remove the correlated measurement residuals so you
   can test this model against the unconstrained model that was estimated at
   point *(a)*. What can you conclude from the comparison of both models, and
   from the model that included the auto-correlations between error terms?

To assume homogeneity of variance for the measurement residuals, we need to
constrain the residuals to be equal across measurement occasions.

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
# Model syntax.
model_ex_1_d <- "
    # Latent intercept variable.
    int =~ 1 * bmi1 + 1 * bmi2 + 1 * bmi3 + 1 * bmi4 + 1 * bmi5 + 1 * bmi6

    # Latent slope variable.
    slp =~ 0 * bmi1 + 1 * bmi2 + 2 * bmi3 + 3 * bmi4 + 4 * bmi5 + 5 * bmi6

    # Latent variances and covariances.
    int ~~ int
    slp ~~ slp
    int ~~ slp

    # Constrained residuals.
    bmi1 ~~ a * bmi1
    bmi2 ~~ a * bmi2
    bmi3 ~~ a * bmi3
    bmi4 ~~ a * bmi4
    bmi5 ~~ a * bmi5
    bmi6 ~~ a * bmi6
"

# Fit model.
model_ex_1_d_fit <- growth(model_ex_1_d, data = data)

# Visualize the model.
semPaths(model_ex_1_d_fit, what = "paths", whatLabels = "std", edge.label.position = 0.7)

# Model summary.
summary(model_ex_1_d_fit, standardized = TRUE, rsquare = TRUE)

# Fit measures.
fitMeasures(model_ex_1_d_fit, fit.measures = fit_indices)
```

To compare the fit of `model_ex_1_d` to `model_ex_1_a` we can take a look at the
fit measures and also perform a *Likelihood Ratio Test* (LRT).

```{r size="footnotesize"}
# Put all fit measures in a data frame or convenience.
fit_measures_ex_1 <- cbind(
    fit_measures_ex_1,
    model_ex_1_d = fitMeasures(model_ex_1_d_fit, fit.measures = fit_indices)
)

# Print all fit measures rounded to four decimals.
round(fit_measures_ex_1, 4)

# Perform the LRT.
anova(model_ex_1_a_fit, model_ex_1_d_fit)
```

For a nuanced explanation, see *Figure 2* that shows the explanation provided by
@newsomLongitudinalStructuralEquation2015 [p. 183] under section *Example 7.1.
Growth Curve Model with Observed Variables*.

\begin{figure}[ht]
    \fbox{\includegraphics[width=0.5\linewidth]{./assets/ex_1_figure_2a.pdf}}
    \fbox{\includegraphics[width=0.5\linewidth]{./assets/ex_1_figure_2b.pdf}}
    \caption{Explanation for \textit{Exercise 1}.}
    \label{fig2}
\end{figure}

\newpage

## Exercise 2

In this exercise you are going to keep using the `health.dat` and extend the
model from *Exercise 1* with a time-invariant covariate, namely the age of the
participants.

a. Compute a new variable `age_c` which is the grand mean centered age of
   the participants.

    - *Tip.* You can obtain the sample mean using `mean(data$age)`.

To create `age_c`, we need to subtract the grand mean from the `age` variable.

```{r size="footnotesize"}
# Create the centered age variable.
data$age_c <- data$age - mean(data$age)
```

b. Estimate the conditional *LGC* model of *BMI* by regressing the latent
   intercept and the latent slope on `age_c`.

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
# Model syntax.
model_ex_2_b <- "
    # Latent intercept variable.
    int =~ 1 * bmi1 + 1 * bmi2 + 1 * bmi3 + 1 * bmi4 + 1 * bmi5 + 1 * bmi6

    # Latent slope variable.
    slp =~ 0 * bmi1 + 1 * bmi2 + 2 * bmi3 + 3 * bmi4 + 4 * bmi5 + 5 * bmi6

    # Latent variances and covariances.
    int ~~ int
    slp ~~ slp
    int ~~ slp

    # Regression equations.
    int + slp ~ age_c
"

# Fit model.
model_ex_2_b_fit <- growth(model_ex_2_b, data = data)

# Visualize the model.
semPaths(model_ex_2_b_fit, what = "paths", whatLabels = "std", edge.label.position = 0.65)

# Model summary.
summary(model_ex_2_b_fit, standardized = TRUE, rsquare = TRUE)

# Fit measures.
fitMeasures(model_ex_2_b_fit, fit.measures = fit_indices)
```

c. Evaluate the fit of this model, and the effect of `age_c` on the latent
   intercept and latent slope.

```{r size="footnotesize"}
# Fit measures.
fitMeasures(model_ex_2_b_fit, fit.measures = fit_indices)
```

d. Compute a new binary age variable called `age_b` for implementing the *MIMIC*
   approach discussed during *Lecture 8*, using the following `R` code:

First we compute the binary variable.

```{r size="footnotesize"}
data$age_b <- ifelse(data$age <= 65, 0, 1)
```

e. Estimate the *MIMIC* model by regressing the intercept and the slope factors
   on the binary age variable. Interpret the effects of the binary age variable.

Now we can implement the *MIMIC* approach. Note that the only thing that changes
is that in our regression equations we now use the binary variable `age_b`
instead of the mean-centered variable `age_c`.

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
# Model syntax.
model_ex_2_e <- "
    # Latent intercept variable.
    int =~ 1 * bmi1 + 1 * bmi2 + 1 * bmi3 + 1 * bmi4 + 1 * bmi5 + 1 * bmi6

    # Latent slope variable.
    slp =~ 0 * bmi1 + 1 * bmi2 + 2 * bmi3 + 3 * bmi4 + 4 * bmi5 + 5 * bmi6

    # Latent variances and covariances.
    int ~~ int
    slp ~~ slp
    int ~~ slp

    # Regression equations.
    int + slp ~ age_b
"

# Fit model.
model_ex_2_e_fit <- growth(model_ex_2_e, data = data)

# Visualize the model.
semPaths(model_ex_2_e_fit, what = "paths", whatLabels = "std", edge.label.position = 0.65)

# Model summary.
summary(model_ex_2_e_fit, standardized = TRUE, rsquare = TRUE)

# Fit measures.
fitMeasures(model_ex_2_e_fit, fit.measures = fit_indices)
```

f. Test the same hypothesis of age differences in trajectories, using the
   multi-group approach and comparing those under $65$ to those aged $65$ and
   older.

For this model, the intercept and slope factor variances were constrained to be
equal in both age groups. The constraints are placed to obtain a more stable
estimate of the variances in $> 65$ group (i.e., which had a small sample size),
and to provide more comparable results to the *MIMIC* modeling approach. The
variance constraints did not significantly degrade the fit of the model, so
these appeared to be empirically justifiable assumptions.

First, we fit a model (i.e., `model_ex_1_a`) where we constrain the latent
variable variances to be equal across the age groups. We can use the `group` and
`group.equal` arguments in `lavaan` to do so. Note that we do not yet constrain
the means of the slope latent variable yet.

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
model_ex_2_f <- "
    # Latent intercept variable.
    int =~ 1 * bmi1 + 1 * bmi2 + 1 * bmi3 + 1 * bmi4 + 1 * bmi5 + 1 * bmi6

    # Latent slope variable.
    slp =~ 0 * bmi1 + 1 * bmi2 + 2 * bmi3 + 3 * bmi4 + 4 * bmi5 + 5 * bmi6

    # Latent variances and covariances.
    int ~~ int
    slp ~~ slp
    int ~~ slp
"

# Fit model.
model_ex_2_f_fit <- growth(model_ex_2_f, data = data, group = "age_b", group.equal = "lv.variances")

# Visualize the model.
semPaths(model_ex_2_f_fit, what = "paths", whatLabels = "std", edge.label.position = 0.7)

# Model summary.
summary(model_ex_2_f_fit, standardized = TRUE, rsquare = TRUE)

# Fit measures.
fitMeasures(model_ex_2_f_fit, fit.measures = fit_indices)
```

Now, on top of the variance constraints, we also constrain the intercept of the
slope latent variable to be equal across groups. Note that since we use the
`group` argument in `lavaan`, this time we have to specify the constraints in
the form of a vector (i.e., `c(a, a)`, which means that both the $< 65$ and
$> 65$ age groups will be applied the same label `a` for the latent slope
variable intercept).

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
model_ex_2_f_slp_con <- "
    # Latent intercept variable.
    int =~ 1 * bmi1 + 1 * bmi2 + 1 * bmi3 + 1 * bmi4 + 1 * bmi5 + 1 * bmi6

    # Latent slope variable.
    slp =~ 0 * bmi1 + 1 * bmi2 + 2 * bmi3 + 3 * bmi4 + 4 * bmi5 + 5 * bmi6

    # Latent variances and covariances.
    int ~~ int
    slp ~~ slp
    int ~~ slp

    # Intercept constraint.
    slp ~ c(a, a) * 1
"

# Fit model.
model_ex_2_f_slp_con_fit <- growth(
    model = model_ex_2_f_slp_con,
    data = data,
    group = "age_b",
    group.equal = "lv.variances"
)

# Visualize the model.
semPaths(model_ex_2_f_slp_con_fit, what = "paths", whatLabels = "std", edge.label.position = 0.7)

# Model summary.
summary(model_ex_2_f_slp_con_fit, standardized = TRUE, rsquare = TRUE)

# Fit measures.
fitMeasures(model_ex_2_f_slp_con_fit, fit.measures = fit_indices)
```

We can now compare the fit of the two models (i.e., `model_ex_2_f` and
`model_ex_2_f_slp_con`) and also perform a *LRT*.

```{r size="footnotesize"}
# Put all fit measures in a data frame or convenience.
fit_measures_ex_2_f <- data.frame(
    model_ex_2_f = fitMeasures(model_ex_2_f_fit, fit.measures = fit_indices),
    model_ex_2_f_slp_con = fitMeasures(model_ex_2_f_slp_con_fit, fit.measures = fit_indices)
)

# Print all fit measures rounded to four decimals.
round(fit_measures_ex_2_f, 4)

# Perform the LRT.
anova(model_ex_2_f_fit, model_ex_2_f_slp_con_fit)
```

\newpage

# References
