---
title: Structural Equation Modeling
author: P.13 - Common Method Bias
date: "`r format(Sys.time(), '%B %d, %Y (%H:%M:%S)')`"
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: ../libs/literature.bib
csl: ../libs/apa.csl
link-citations: yes
colorlinks: true
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
  - \usepackage{longtable}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Set the path only for interactive sessions.
if (interactive()) setwd("./Practical 13 - Common Method Bias")
```

---

# Lab Description

In this assignment you will learn how to fit *Structural Equation Models* (SEM)
while accounting for *Common Method Bias*. For this practical you will need the
following packages: `lavaan`, `semPlot`, and `haven`. You can install and load
these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot", "haven"))

# Load the packages.
library(lavaan)
library(semPlot)
library(haven)
```

Please note that the data for following assignment are kindly provided by [dr.
Pieter de Rooij](https://pure.buas.nl/en/persons/pieter-de-rooij) from *Breda
University of Applied Sciences*. We acquired permission to use the data in the
course *Research Master: Structural Equation Modeling and Analysis of
Longitudinal Data*. Should you want to use the data outside the scope of this
course, please make sure to obtain written approval from dr. Pieter de Rooij (at
rooy.h@buas.nl).

## More about the data

This dataset contains $27$ items that seek to measure $9$ dimensions (i.e.,
constructs) of visitors’ experiences when they attend a performance in the
performing arts sector (e.g., drama, dance, musical, stand up comedy, opera,
classical music). Each construct is indicated by $3$ items as shown in *Table
1*. Scales were rated on $5$ point *Likert* scales (i.e. from $1 = \text{totally
disagree}$ to $5 = \text{totally agree}$).

\newpage

\begin{longtable}{llp{11cm}}
    \caption{Latent constructs and manifest variables in \texttt{data.sav}.} \\
    \label{tab:1} \\

    \hline \multicolumn{1}{l}{\textbf{Dimensions}} & \multicolumn{1}{l}{\textbf{Variable}} & \multicolumn{1}{l}{\textbf{Item text}} \\ \hline
    \endfirsthead

    \multicolumn{3}{r}%
    {{\bfseries \tablename\ \thetable{} -- continued}} \\
    \hline \multicolumn{1}{l}{\textbf{Dimensions}} & \multicolumn{1}{l}{\textbf{Variable}} & \multicolumn{1}{l}{\textbf{Item text}} \\ \hline
    \endhead

    \hline \multicolumn{3}{r}{{Continued on next page}}
    \endfoot

    \hline
    \endlastfoot

    Artistic value          & \texttt{artval1}   & The concert/ show had artistic value. \\
                            & \texttt{artval2}   & The concert/ show was a form of art. \\
                            & \texttt{artval3}   & The concert/ show was of a high artistic level. \\
    \hline
    Beauty                  & \texttt{beauty1}   & The concert/ how was beautiful. \\
                            & \texttt{beauty2}   & I enjoyed the concert/ show. \\
                            & \texttt{beauty3}   & The concert/ how was a beautiful experience. \\
    \hline
    Cultural relaxation     & \texttt{relax1}    & Through my visit to the concert/ show I had a nice evening out. \\
                            & \texttt{relax2}    & Through my visit to the concert/ show I had the feeling I was away from it all. \\
                            & \texttt{relax3}    & Through my visit to the concert/ show I have been able to relax. \\
    \hline
    Cultural stimulation    & \texttt{stim1}     & Through my visit to the concert/ show I got food for thought. \\
                            & \texttt{stim2}     & Through my visit to the concert/ show I felt intellectually stimulated. \\
                            & \texttt{stim3}     & Through my visit to the concert/ show I have been challenged to think about certain things. \\
    \hline
    Cultural transmission   & \texttt{trans1}    & Through my visit to the concert/ show I transmitted my cultural interests to important others (children, grandchildren, family, friends.) \\
                            & \texttt{trans2}    & Through my visit to the concert/ show I have shared my cultural interests with important others (children, grandchildren, family, friends.) \\
                            & \texttt{trans3}    & Through my visit to the concert/ show I have brought people that are important to me into contact with this form of art. \\
    \hline
    Social attraction       & \texttt{attr1}     & During my visit to the concert/ show I was with people that like the same things as I do. \\
                            & \texttt{attr2}     & During my visit to the concert/ show I had the feeling I was with like-minded people. \\
                            & \texttt{attr3}     & During my visit to the concert/ show I was with people that have similar interests as me. \\
    \hline
    Social bonding          & \texttt{bond1}     & My visit to the concert/ show was a nice opportunity to be together with family and friends. \\
                            & \texttt{bond2}     & Through my visit to the concert/ show I had a nice evening out with family or friends. \\
                            & \texttt{bond3}     & Through my visit to the concert/ show I have been able to strengthen the bonds with family or friends. \\
    \hline
    Social distinction      & \texttt{dist1}     & My visit to the concert/ show is a good topic for conversation at drinks or other social activities. \\
                            & \texttt{dist2}     & I like to tell other people how I experienced the concert/ show. \\
                            & \texttt{dist3}     & People in my social environment appreciate a visit to this concert/ show. \\
    \hline
    Social duty             & \texttt{duty1}     & Through my visit to the concert/ show I spent a pleasant time with colleagues or business relations. \\
                            & \texttt{duty2}     & Through my visit to the concert/ show I met interesting people from my professional network. \\
                            & \texttt{duty3}     & Through my visit to the concert/ show I strengthened my professional network. \\
\end{longtable}

Start by loading the *SPSS* dataset `data.sav` into `R` using the `haven`
package.

- *Hint.* Check out the `read_spss` function from the `R` package `haven`.

## Questions

1. Estimate a *CFA* model for the $9$ constructs. Evaluate the fit of this
   model.

2. Estimate a *CFA* model for the $9$ constructs together with a *method factor*
   to control for potential *common method bias*. Model the *common method
   variance* with a single *latent method variable*. Implement equality
   constraints on the loadings of the method factor and also implement the
   necessary constraints for the associations between method and content
   factors.

3. Test both models against each other using the *Likelihood Ratio Test* (LRT).
   What do you conclude?

4. Do you see other possibilities to improve the fit of the model?
