---
title: Structural Equation Modeling
author: P.13 - Common Method Bias
date: "`r format(Sys.time(), '%B %d, %Y (%H:%M:%S)')`"
output:
    pdf_document:
        highlight: zenburn
citation_package: biblatex
bibliography: ../libs/literature.bib
csl: ../libs/apa.csl
link-citations: yes
colorlinks: true
header-includes:
  - \usepackage{leading}
  - \usepackage{subcaption}
  - \usepackage{csquotes}
  - \usepackage{bm}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{hyperref}
  - \leading{16pt}
  - \newcommand{\mat}[1]{\bm{#1}}
  - \renewcommand{\vec}[1]{\bm{#1}}
  - \usepackage{longtable}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Output width.
options(width = 100)

# Reduce code chunk and output size.
# See: https://stackoverflow.com/a/46526740/5252007.
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

# Load libraries.
library(lavaan)
library(semPlot)
library(haven)

# Set the path only for interactive sessions.
if (interactive()) setwd("./Practical 13 - Common Method Bias")

# Load the data.
data <- read_spss("./data/data.sav")
```

---

# Lab Description

In this assignment you will learn how to fit *Structural Equation Models* (SEM)
while accounting for *Common Method Bias*. For this practical you will need the
following packages: `lavaan`, `semPlot`, and `haven`. You can install and load
these packages using the following code:

```{r size="footnotesize", eval=FALSE}
# Install packages.
install.packages(c("lavaan", "semPlot", "haven"))

# Load the packages.
library(lavaan)
library(semPlot)
library(haven)
```

Please note that the data for following assignment are kindly provided by [dr.
Pieter de Rooij](https://pure.buas.nl/en/persons/pieter-de-rooij) from *Breda
University of Applied Sciences*. We acquired permission to use the data in the
course *Research Master: Structural Equation Modeling and Analysis of
Longitudinal Data*. Should you want to use the data outside the scope of this
course, please make sure to obtain written approval from dr. Pieter de Rooij (at
rooy.h@buas.nl).

## More about the data

This dataset contains $27$ items that seek to measure $9$ dimensions (i.e.,
constructs) of visitors’ experiences when they attend a performance in the
performing arts sector (e.g., drama, dance, musical, stand up comedy, opera,
classical music). Each construct is indicated by $3$ items as shown in *Table
1*. Scales were rated on $5$ point *Likert* scales (i.e. from $1 = \text{totally
disagree}$ to $5 = \text{totally agree}$).

\newpage

\begin{longtable}{llp{11cm}}
    \caption{Latent constructs and manifest variables in \texttt{data.sav}.} \\
    \label{tab:1} \\

    \hline \multicolumn{1}{l}{\textbf{Dimensions}} & \multicolumn{1}{l}{\textbf{Variable}} & \multicolumn{1}{l}{\textbf{Item text}} \\ \hline
    \endfirsthead

    \multicolumn{3}{r}%
    {{\bfseries \tablename\ \thetable{} -- continued}} \\
    \hline \multicolumn{1}{l}{\textbf{Dimensions}} & \multicolumn{1}{l}{\textbf{Variable}} & \multicolumn{1}{l}{\textbf{Item text}} \\ \hline
    \endhead

    \hline \multicolumn{3}{r}{{Continued on next page}}
    \endfoot

    \hline
    \endlastfoot

    Artistic value          & \texttt{artval1}   & The concert/ show had artistic value. \\
                            & \texttt{artval2}   & The concert/ show was a form of art. \\
                            & \texttt{artval3}   & The concert/ show was of a high artistic level. \\
    \hline
    Beauty                  & \texttt{beauty1}   & The concert/ how was beautiful. \\
                            & \texttt{beauty2}   & I enjoyed the concert/ show. \\
                            & \texttt{beauty3}   & The concert/ how was a beautiful experience. \\
    \hline
    Cultural relaxation     & \texttt{relax1}    & Through my visit to the concert/ show I had a nice evening out. \\
                            & \texttt{relax2}    & Through my visit to the concert/ show I had the feeling I was away from it all. \\
                            & \texttt{relax3}    & Through my visit to the concert/ show I have been able to relax. \\
    \hline
    Cultural stimulation    & \texttt{stim1}     & Through my visit to the concert/ show I got food for thought. \\
                            & \texttt{stim2}     & Through my visit to the concert/ show I felt intellectually stimulated. \\
                            & \texttt{stim3}     & Through my visit to the concert/ show I have been challenged to think about certain things. \\
    \hline
    Cultural transmission   & \texttt{trans1}    & Through my visit to the concert/ show I transmitted my cultural interests to important others (children, grandchildren, family, friends.) \\
                            & \texttt{trans2}    & Through my visit to the concert/ show I have shared my cultural interests with important others (children, grandchildren, family, friends.) \\
                            & \texttt{trans3}    & Through my visit to the concert/ show I have brought people that are important to me into contact with this form of art. \\
    \hline
    Social attraction       & \texttt{attr1}     & During my visit to the concert/ show I was with people that like the same things as I do. \\
                            & \texttt{attr2}     & During my visit to the concert/ show I had the feeling I was with like-minded people. \\
                            & \texttt{attr3}     & During my visit to the concert/ show I was with people that have similar interests as me. \\
    \hline
    Social bonding          & \texttt{bond1}     & My visit to the concert/ show was a nice opportunity to be together with family and friends. \\
                            & \texttt{bond2}     & Through my visit to the concert/ show I had a nice evening out with family or friends. \\
                            & \texttt{bond3}     & Through my visit to the concert/ show I have been able to strengthen the bonds with family or friends. \\
    \hline
    Social distinction      & \texttt{dist1}     & My visit to the concert/ show is a good topic for conversation at drinks or other social activities. \\
                            & \texttt{dist2}     & I like to tell other people how I experienced the concert/ show. \\
                            & \texttt{dist3}     & People in my social environment appreciate a visit to this concert/ show. \\
    \hline
    Social duty             & \texttt{duty1}     & Through my visit to the concert/ show I spent a pleasant time with colleagues or business relations. \\
                            & \texttt{duty2}     & Through my visit to the concert/ show I met interesting people from my professional network. \\
                            & \texttt{duty3}     & Through my visit to the concert/ show I strengthened my professional network. \\
\end{longtable}

## Questions

Start by loading the *SPSS* dataset `data.sav` into `R` using the `haven`
package.

- *Hint.* Check out the `read_spss` function from the `R` package `haven`.

Set the working directory to the location where your data file has been
downloaded and load the data.

```{r size="footnotesize", eval=FALSE}
# For example.
setwd("/Users/mihai/Downloads")

# Load data.
data <- read_spss("data.sav")

# Inspect the data.
View(data)
```

List the variables.

```{r size="footnotesize", eval=FALSE}
# List variables.
str(data)
```

Specify which fit measures we are interested in:

```{r size='scriptsize'}
# Fit indices to print.
fit_indices <- c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", "rmsea.pvalue", "srmr")
```

1. Estimate a *CFA* model for the $9$ constructs. Evaluate the fit of this
   model.

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
# Model syntax.
model_q_1 <- "
    # Measurement part.
    artval  =~ exp_artval1 + exp_artval2 + exp_artval3
    attr    =~ exp_attr1   + exp_attr2   + exp_attr3
    beaty   =~ exp_beauty1 + exp_beauty2 + exp_beauty3
    bond    =~ exp_bond1   + exp_bond2   + exp_bond3
    dist    =~ exp_dist1   + exp_dist2   + exp_dist3
    duty    =~ exp_duty1   + exp_duty2   + exp_duty3
    esc     =~ exp_esc1    + exp_esc2    + exp_esc3
    stim    =~ exp_stim1   + exp_stim2   + exp_stim3
    trans   =~ exp_trans1  + exp_trans2  + exp_trans3
"

# Fit model.
model_q_1_fit <- cfa(model_q_1, data = data)

# Visualize the model.
semPaths(
    model_q_1_fit,
    what = "path",
    whatLabels = "omit",
    sizeMan = 6.5,
    sizeLat = 6,
    style = "lisrel",
    layout = "spring",
    mar = c(1.5, 1.5, 1.5, 1.5)
)

# Model summary.
summary(model_q_1_fit, standardized = TRUE)

# Fit measures.
fitMeasures(model_q_1_fit, fit.measures = fit_indices)
```

2. Estimate a *CFA* model for the $9$ constructs together with a *method factor*
   to control for potential *common method bias*. Model the *common method
   variance* with a single *latent method variable*. Implement equality
   constraints on the loadings of the method factor and also implement the
   necessary constraints for the associations between method and content
   factors.

```{r size="footnotesize", dev="pdf", fig.width=7, fig.height=7, out.width="300px", fig.align="center"}
# Model syntax.
model_q_2 <- "
    # Measurement part.
    artval  =~ exp_artval1 + exp_artval2 + exp_artval3
    attr    =~ exp_attr1   + exp_attr2   + exp_attr3
    beaty   =~ exp_beauty1 + exp_beauty2 + exp_beauty3
    bond    =~ exp_bond1   + exp_bond2   + exp_bond3
    dist    =~ exp_dist1   + exp_dist2   + exp_dist3
    duty    =~ exp_duty1   + exp_duty2   + exp_duty3
    esc     =~ exp_esc1    + exp_esc2    + exp_esc3
    stim    =~ exp_stim1   + exp_stim2   + exp_stim3
    trans   =~ exp_trans1  + exp_trans2  + exp_trans3

    # Common method variance latent variable with constrained loadings.
    cmv =~ NA * exp_artval1 +
            a * exp_artval1 + a * exp_artval2 + a * exp_artval3 +
            a * exp_attr1   + a * exp_attr2   + a * exp_attr3 +
            a * exp_beauty1 + a * exp_beauty2 + a * exp_beauty3 +
            a * exp_bond1   + a * exp_bond2   + a * exp_bond3 +
            a * exp_dist1   + a * exp_dist2   + a * exp_dist3 +
            a * exp_duty1   + a * exp_duty2   + a * exp_duty3 +
            a * exp_esc1    + a * exp_esc2    + a * exp_esc3 +
            a * exp_stim1   + a * exp_stim2   + a * exp_stim3 +
            a * exp_trans1  + a * exp_trans2  + a * exp_trans3

    # Fix variance of the `cmv` latent variable.
    cmv ~~ 1 * cmv

    # Fix covariances between the `cmv` and the other latent variables.
    artval ~~ 0 * cmv
    attr   ~~ 0 * cmv
    beaty  ~~ 0 * cmv
    bond   ~~ 0 * cmv
    dist   ~~ 0 * cmv
    duty   ~~ 0 * cmv
    esc    ~~ 0 * cmv
    stim   ~~ 0 * cmv
    trans  ~~ 0 * cmv
"

# Fit model.
model_q_2_fit <- cfa(model_q_2, data = data)

# Visualize the model.
semPaths(
    model_q_2_fit,
    what = "path",
    whatLabels = "omit",
    sizeMan = 6.5,
    sizeLat = 6,
    style = "lisrel",
    layout = "spring",
    mar = c(1.5, 1.5, 1.5, 1.5)
)
```

```{r size="footnotesize", dev="pdf", fig.width=10, fig.height=6, out.width="450px", fig.align="center"}
# Regular tree layout.
semPaths(
    model_q_2_fit,
    what = "path",
    whatLabels = "omit",
    sizeMan = 3,
    sizeLat = 6,
    style = "lisrel",
    mar = c(1.5, 1.5, 1.5, 1.5)
)

# Model summary.
summary(model_q_2_fit, standardized = TRUE)

# Fit measures.
fitMeasures(model_q_2_fit, fit.measures = fit_indices)
```

3. Test both models against each other using the *Likelihood Ratio Test* (LRT).
   What do you conclude?

We can first compare the fit indices.

```{r size="footnotesize"}
# Put all fit measures together.
fit_measures_all <- data.frame(
    model_q_1 = fitMeasures(model_q_1_fit, fit.measures = fit_indices),
    model_q_2 = fitMeasures(model_q_2_fit, fit.measures = fit_indices)
)

# Print the measures with four decimals.
print(round(fit_measures_all, 4))
```

Now we can also perform a *LRT*.

```{r size="footnotesize"}
# Perform LRT.
anova(model_q_2_fit, model_q_1_fit)
```

We see evidence that including a factor to account for common method bias
improves the model fit. The model estimated at point *(1)* (i.e., `model_q_1`)
does not fit equally well as the more complex model fit at point *(2)* (i.e., `model_q_2`).
We prefer `model_q_2` with the method factor added.

4. Do you see other possibilities to improve the fit of the model?

*What do you think?*
